---
title: "Calculation of 5-year age specific birth rates"
author: "Shelmith Kariuki"
date: "2/8/2021"
output: html_document
---

```{r setup, include=FALSE}
# rm(list = ls())
basedir <- normalizePath("../../ddharmony/")
knitr::opts_knit$set(root.dir = basedir, message=FALSE, warning=FALSE)

```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 2000)
options(scipen = 999)
```

#### 01.) Load the packages that are required.

```{r, message=FALSE, warning=FALSE}

###  create a vector of packages to be installed
pkgs <- c("tidyverse","data.table","DT","lubridate", "readxl","googlesheets4",
          "openxlsx", "kableExtra","zoo", "stringr", "EnvStats", "trend", "beepr",
          "devtools","httr")

###  Check if there are packages you want to load, that are not already installed. 
miss_pkgs <- pkgs[!pkgs %in% installed.packages()[,1]] 

###  Installing the missing packages
if(length(miss_pkgs)>0){
  install.packages(miss_pkgs)
}

###  Loading all the packages
invisible(lapply(pkgs,library,character.only=TRUE))
library(DDSQLtools)

###  Remove the objects that are no longer required
rm(miss_pkgs)
rm(pkgs)

```


#### 02.) Load location data

```{r, message=FALSE, warning=FALSE}
load("Shel/Locations.RData")

```

#### 1.) Load the 5-year population data

```{r, message=FALSE, warning=FALSE}

pop5_df0 <- read_xlsx("Shel/WPP2019_POP_F15_3_ANNUAL_POPULATION_BY_AGE_FEMALE.xlsx")
pop5_df <- pop5_df0[-1:-11, ] 

colnames(pop5_df) <- pop5_df[1,]
pop5_df <- pop5_df[-1, ]

```


#### 2.) Reshape the data so that we have age labels on one column, and population counts in another, and rename the agelabel variable to AgeLabel_denom.

```{r, message=FALSE, warning=FALSE}
pop5_df <- pop5_df %>% 
  mutate(`Country code` = as.numeric(trimws(`Country code`))) %>% 
  gather("AgeLabel_denom", "pop.count", contains("-"), contains("+")) %>% 
  rename(LocID = `Country code`,
         TimeLabel = `Reference date (as of 1 July)`) 

## Deal with scientific notation
pop5_df <- pop5_df %>% 
  mutate(sci_exists = ifelse(pop.count %in% grep("E", pop.count, value = T, ignore.case = FALSE),
                             1, 0),
         sci_not = ifelse(sci_exists == 1, as.numeric(trimws(gsub(".*E", "", pop.count))), NA),
         pop.count = ifelse(sci_exists == 1, trimws(gsub("E.*", "", pop.count)), trimws(pop.count)),
         pop.count = as.numeric(pop.count)) %>% 
  mutate(pop.count = ifelse(sci_exists == 1, pop.count * 10^sci_not, 
                            pop.count)) %>% 
  select(-sci_exists, -sci_not) %>% 
  arrange(TimeLabel)

```


#### 3. Specify the location

```{r, message=FALSE, warning=FALSE}
# lid <- sample(unique(pop5_df$LocID), 1)
lid <- 466
pop5_df2 <- pop5_df %>% 
  filter(LocID == lid)

```


#### 4.) Load births data

```{r, message=FALSE, warning=FALSE}

req <- GET("https://api.github.com/repos/sarahertog/ddharmony/git/trees/main?recursive=1")
stop_for_status(req)
filelist <- unlist(lapply(content(req)$tree, "[", "path"), use.names = F)

for (filename in filelist) {
  
  one_function <- paste0("https://github.com/sarahertog/ddharmony/blob/main/", filename, "?raw=TRUE")
  source_url(one_function)
  rm(one_function)
}
rm(req, filelist, filename)

# Births by age of mother: vr
births_vr <- try(DDharmonize_validate_BirthCounts(locid = lid, process = "vr", times = 1950:2020, retainKeys = TRUE))

# Births by age of mother: census
births_census <- try(DDharmonize_validate_BirthCounts(locid = lid, process = "census", times = 1950:2020, retainKeys = TRUE))

# if(any(class(births_vr) %in% "data.frame") & any(class(births_census) %in% "data.frame")){
#   
#     common = names(births_vr)[names(births_vr) %in% names(births_census)]
#   births_census[common] <- lapply(common, function(x) {
#     match.fun(paste0("as.", class(births_vr[[x]])))(births_census[[x]])})
#   
#   births_df <- bind_rows(births_vr,births_census )
# }else
#  if(any(class(births_vr) %in% "data.frame") & any(!class(births_census) %in% "data.frame" )){
#   births_df <- births_vr
# }else 
#  if(any(!class(births_vr) %in% "data.frame") & any(class(births_census) %in% "data.frame" )){  births_df <- births_census
#  }  
births_df <- births_census

```


#### 5.) Merge the births data with the Location data

```{r, message=FALSE, warning=FALSE}
births_df2 <- Locations %>% 
  right_join(.,births_df , by = c("LocID", "LocName")) %>% 
  rename(AgeLabel_num = AgeLabel)

```


#### 6.) Extract the 5-year counts and rename the agelabel variable to AgeLabel_num.

```{r, message=FALSE, warning=FALSE}
births_df2_5 <- births_df2 %>% 
  filter(five_year == TRUE)

```

#### 7.) Generate a variable that shows the maximum age label

```{r, message=FALSE, warning=FALSE}
if(nrow(births_df2_5) >0){

births_df2_5 <- births_df2_5 %>% 
  separate(AgeLabel_num, into = c("ll", "ul"), sep = "-", remove = FALSE) %>% 
  mutate(ul = ifelse(is.na(ul) & 
                       AgeLabel_num %in% grep("\\d+", AgeLabel_num, value = TRUE), AgeLabel_num, ul)) %>% 
  mutate(ul = as.numeric(trimws(gsub("\\+", "", ul)))) %>% 
  group_by(id) %>% 
  mutate(max_age = ifelse(any(ul == max(ul, na.rm = TRUE)), 
                          AgeLabel_num[ul == max(ul, na.rm = TRUE)], NA)) %>% 
  ungroup() 
}

births_df2_5 <- births_df2_5 %>% 
  mutate(TimeLabel = as.character(TimeLabel))
  
```


#### 8.) Merge the population data with the births data.

```{r, message=FALSE, warning=FALSE}
merged_df <- pop5_df2 %>% 
  #mutate(LocID = as.numeric(trimws(LocID))) %>% 
  #        TimeLabel = as.numeric(trimws(TimeLabel))) %>% 
  full_join(.,births_df2_5 , by = c("LocID", "TimeLabel",  "AgeLabel_denom" = "AgeLabel_num")) %>% 
  rename(births.count = DataValue, 
         AgeLabel = AgeLabel_denom) %>% 
  relocate(c(TimeLabel, AgeLabel, pop.count, births.count), .after = last_col()) %>% 
  arrange(TimeLabel)

if(!"max_age" %in% names(merged_df)){
merged_df$max_age <- NA
}

## Fill in the missing values
merged_df <- merged_df %>% 
  group_by(TimeLabel) %>% 
  mutate(id = na.locf0(id),
         max_age = na.locf0(max_age, fromLast = TRUE)) %>% 
  ungroup() %>% 
  group_by(id, TimeLabel) %>% 
  mutate(max_age = ifelse(all(is.na(births.count)), NA, max_age)) %>% 
  ungroup()

## Split age label into ll and ul. 
merged_df <- merged_df %>% 
  separate(AgeLabel, into = c("ll", "ul"), sep = "-", remove = FALSE) %>% 
  mutate(ul = ifelse(is.na(ul) & 
                       AgeLabel %in% grep("\\d+", AgeLabel, value = TRUE), AgeLabel, ul)) %>% 
  mutate(max_age2 = max_age) %>% 
  mutate(across(c(max_age2, ll, ul), ~as.numeric(trimws(gsub("\\+", "", .))))) %>% 
  mutate(juncture = ifelse(ul >= max_age2, 1, 0)) %>% 
  mutate(AgeLabel = ifelse(is.na(juncture), AgeLabel,
           ifelse(juncture == 1 & AgeLabel!= "Total"  , max_age, 
                  AgeLabel ))) %>% 
  group_by(id) %>% 
  mutate(pop.count = ifelse(is.na(juncture), pop.count,
           ifelse(juncture == 1, sum(pop.count[juncture == 1], na.rm = TRUE),
                            pop.count))) %>% 
  ungroup()

## Drop the ages that have been collapsed into a smaller bracket
merged_df <- merged_df %>% 
  mutate(todrop = ifelse(AgeLabel == max_age & is.na(births.count) & !is.na(pop.count),
                         "drop", "keep")) %>% 
  filter(todrop == "keep"|is.na(todrop)) %>% 
  group_by(id) %>% 
  mutate(pop.count = ifelse(is.na(pop.count), sum(pop.count, na.rm = TRUE), pop.count)) %>% 
  select(-ll, -ul, -max_age, -juncture, -todrop, -max_age2) %>% 
  ungroup()

## Generate a variable that shows whether births are missing, and another that shows whether population counts are missing
merged_df <- merged_df %>% 
    group_by(id, TimeLabel) %>% 
  mutate(missing_births = ifelse(all(is.na(births.count)|all(births.count == 0)), TRUE, FALSE),
         missing_popcounts = ifelse(all(is.na(pop.count)|all(pop.count == 0)), TRUE, FALSE)) %>% 
  ungroup()

```


#### 9.) Calculate the age specific birth rates for each of the years and age labels.

```{r, message=FALSE, warning=FALSE}
merged_df <- merged_df %>% 
  mutate(asbr = births.count / pop.count)

```


#### 10.) Reshape the data so that we have category on one variable (pop.count, births.count, asbr).

```{r, message=FALSE, warning=FALSE}
merged_df2 <- merged_df %>% 
  select(-Index, -AgeStart, -AgeEnd, -AgeSpan, -AgeSort, -abridged, -complete, 
         -non_standard) %>% 
  mutate(pop.count = ifelse(is.na(pop.count), 0, pop.count),
         births.count = ifelse(is.na(births.count), 0, births.count),
         asbr = ifelse(is.na(asbr), 0, asbr)) %>% 
  group_by(id, TimeLabel) %>% 
  mutate_all(~na.locf0(., fromLast = FALSE)) %>% 
  ungroup() %>% 
  gather("category", "value", pop.count, births.count, asbr) %>%
  mutate(AgeLabel = trimws(str_squish(AgeLabel))) %>% 
  group_by(id, TimeLabel) %>% 
  spread(AgeLabel, value) %>% 
  arrange(TimeLabel) %>% 
  ungroup()
  
if(!"Total" %in% names(merged_df2)){
  merged_df2$Total <- NA
}
  
  vars <- c("0-4", "5-9", "10-14", "15-19", "20-24", "25-29", "30+" , "30-34","35-39","40+", "40-44", "45+", "45-49", "50+", "50-54", "55+", "55-59","60+",  "60-64", "65-69", "70-74", "75-79", "80-84", "85-89", "90-94", "95-99", "100+")
  
merged_df2 <- merged_df2 %>% 
  select(Variant: category, any_of(vars) , 
         everything()) 
  
agelabel_vars <- merged_df2 %>% 
  select(-Variant: -category, -Total) %>% 
  names()


merged_df2$Total <- ifelse(is.na(merged_df2$Total), 
                           apply(merged_df2[,agelabel_vars], 1, function(x) 
                             sum(x, na.rm = TRUE)),merged_df2$Total)

vars <- grep("\\d", names(merged_df2), value = TRUE)
vars <- vars[!vars %in% "ISO3Alpha"]

merged_df2 <- merged_df2 %>% 
  select(id, TimeLabel:category, any_of(vars) ,  Total)

```


#### 11.) Extract the estimates.

```{r, }
asbr5_df <- merged_df2 %>% 
  filter(category == "asbr")
```

```{r}
# 234, 292, 500, 584 missing from pop data.
```

